---
title: "CAPM Model"
author: "Rodrigo Kang"

format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-line-numbers: true
    code-overflow: wrap
    theme: cosmo
    embed-resources: true

execute:
  warning: false
  message: false

editor: visual
---

## Overview

This notebook accompanies the project **CAPM and Fama–French Models** and contains the computational implementation supporting the methodology, experiments, and results discussed in the portfolio chapter.

The material presented here focuses on practical implementation details, including data preparation, model development, training procedures, evaluation workflows, and complementary computational experiments.

## Imports and Configuration

This section initializes the main libraries used throughout the notebook.

The analysis uses historical market data obtained from Yahoo Finance together with standard R libraries for data manipulation, visualization, and linear regression modeling.

The goal is to evaluate a simple CAPM specification by comparing the returns of an individual asset against the S&P 500 index as a proxy for the overall market.

```{r}
Sys.setenv(LANGUAGE = "en")

library(quantmod)
library(dplyr)
library(ggplot2)
library(broom)
library(PerformanceAnalytics)
library(gt)
```

```{r}
outputs_dir <- "1-capm-model-figures"

dir.create(outputs_dir, showWarnings = FALSE)
```

## Asset Selection and Market Benchmark

Apple Inc. (`AAPL`) is used as the individual asset, while the S&P 500 index (`^GSPC`) is used as a proxy for the market portfolio.

The analysis uses five years of daily adjusted closing prices obtained from Yahoo Finance.

```{r}
asset_ticker <- "AAPL"
market_ticker <- "^GSPC"

start_date <- "2020-01-01"
end_date <- "2025-01-01"
```

## Data Collection

Daily closing prices for both the selected asset and the market index are downloaded from Yahoo Finance and combined into a single dataset for further analysis.

The downloaded prices are adjusted for corporate actions when available, making them suitable for return calculation.

```{r}
asset_data <- getSymbols(
  asset_ticker,
  src = "yahoo",
  from = start_date,
  to = end_date,
  auto.assign = FALSE
)

market_data <- getSymbols(
  market_ticker,
  src = "yahoo",
  from = start_date,
  to = end_date,
  auto.assign = FALSE
)

asset_prices <- Ad(asset_data)
market_prices <- Ad(market_data)

prices <- merge(asset_prices, market_prices)

colnames(prices) <- c("asset", "market")

prices <- na.omit(prices)

head(prices)
```

## Price Series Visualization

The price series are inspected before computing returns. Since the asset and the market index are expressed on different price scales, the visualization is mainly used as a preliminary inspection of the downloaded data.

```{r}
prices_df <- data.frame(
  date = index(prices),
  coredata(prices)
)

closing_prices_plot <- ggplot(prices_df, aes(x = date)) +
  geom_line(aes(y = asset, colour = "Asset")) +
  geom_line(aes(y = market, colour = "Market")) +
  labs(
    title = "Closing Prices",
    x = "Date",
    y = "Price",
    colour = NULL
  ) +
  theme_minimal()

closing_prices_plot
```

```{r}
ggsave(
  filename = file.path(outputs_dir, "closing-prices.png"),
  plot = closing_prices_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```

## Return Calculation

Daily returns are computed from the adjusted closing price series and used as the main variables in the CAPM regression analysis.

These returns provide a normalized representation of market behavior and allow the linear relationship between the asset and the market portfolio to be estimated.

```{r}
returns <- na.omit(Return.calculate(prices, method = "discrete"))

returns_df <- data.frame(
  date = index(returns),
  coredata(returns)
)

head(returns_df)
```

## Exploratory Analysis

The relationship between asset and market returns is briefly explored before estimating the regression model.

The scatter plot provides an initial visual assessment of the linear association assumed by the CAPM framework.

```{r}
daily_returns_plot <- ggplot(returns_df, aes(x = market, y = asset)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Asset vs Market Daily Returns",
    x = "Market Returns",
    y = "Asset Returns"
  ) +
  theme_minimal()

daily_returns_plot
```

```{r}
ggsave(
  filename = file.path(outputs_dir, "daily-returns-scatter.png"),
  plot = daily_returns_plot,
  width = 7,
  height = 5,
  dpi = 300
)
```

## CAPM Estimation

A simple linear regression is estimated using market returns as the independent variable and asset returns as the dependent variable.

The estimated coefficients are interpreted as the asset alpha and beta under the CAPM specification.

```{r}
capm_model <- lm(asset ~ market, data = returns_df)

capm_coefficients_table <- tidy(capm_model)
```

```{r}
capm_coefficients_gt <- capm_coefficients_table |>
  mutate(
    estimate = round(estimate, 6),
    std.error = round(std.error, 6),
    statistic = round(statistic, 4),
    p.value = round(p.value, 6)
  ) |>
  gt() |>
  tab_header(
    title = "CAPM Regression Coefficients"
  ) |>
  cols_label(
    term = "Term",
    estimate = "Estimate",
    std.error = "Std. Error",
    statistic = "t-statistic",
    p.value = "p-value"
  )

capm_coefficients_gt
```

```{r}
gtsave(
  data = capm_coefficients_gt,
  filename = file.path(outputs_dir, "capm-coefficients-table.html")
)
```

```{r}
capm_diagnostics_table <- tibble(
  Metric = c(
    "R-squared",
    "Adjusted R-squared",
    "F-statistic",
    "Observations",
    "AIC",
    "BIC"
  ),
  Value = c(
    summary(capm_model)$r.squared,
    summary(capm_model)$adj.r.squared,
    summary(capm_model)$fstatistic[1],
    nobs(capm_model),
    AIC(capm_model),
    BIC(capm_model)
  )
)
```

```{r}
capm_diagnostics_gt <- capm_diagnostics_table |>
  mutate(
    Value = round(Value, 4)
  ) |>
  gt() |>
  tab_header(
    title = "CAPM Model Diagnostics"
  )

capm_diagnostics_gt
```

```{r}
gtsave(
  data = capm_diagnostics_gt,
  filename = file.path(outputs_dir, "capm-model-diagnostics.html")
)
```

## Model Evaluation

The estimated regression coefficients are inspected together with the fitted relationship between asset and market returns.

The regression line provides a visual representation of the sensitivity of the selected asset to movements in the market portfolio.

```{r}
alpha <- coef(capm_model)[["(Intercept)"]]
beta <- coef(capm_model)[["market"]]

cat("Alpha:", round(alpha, 6), "\n")
cat("Beta:", round(beta, 4), "\n")
```

```{r}
capm_regression_plot <- ggplot(returns_df, aes(x = market, y = asset)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, colour = "red") +
  labs(
    title = "CAPM Regression",
    x = "Market Returns",
    y = "Asset Returns"
  ) +
  theme_minimal()

capm_regression_plot
```

```{r}
ggsave(
  filename = file.path(outputs_dir, "capm-regression.png"),
  plot = capm_regression_plot,
  width = 7,
  height = 5,
  dpi = 300
)
```
