---
title: "CAPM Model"
author: "Rodrigo Kang"

format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-line-numbers: true
    code-overflow: wrap
    theme: cosmo
    embed-resources: true

editor: visual
---

## Overview

This notebook accompanies the project [**CAPM and Fama–French Models**](https://rodrigokang.github.io/portfolio/1-quantitative-finance/1-capm-and-fama-french-factor-model.html) and contains the computational implementation supporting the methodology, experiments, and results discussed in the portfolio chapter.

The material presented here focuses on practical implementation details, including data preparation, model development, training procedures, evaluation workflows, and complementary computational experiments.

## Imports and Configuration

This notebook contains the R implementation accompanying the CAPM analysis. The workflow uses historical market data from Yahoo Finance and focuses on the practical steps required to estimate a simple linear relationship between an individual asset and the market.

```{r}
library(quantmod)
library(dplyr)
library(ggplot2)
library(broom)
library(PerformanceAnalytics)
```

## Asset Selection and Market Benchmark

Apple Inc. (`AAPL`) is used as the individual asset, while the S&P 500 index (`^GSPC`) is used as a proxy for the market portfolio.

The analysis uses five years of daily market data.

```{r}
asset_ticker <- "AAPL"
market_ticker <- "^GSPC"

start_date <- "2020-01-01"
end_date <- "2025-01-01"
```

## Data Collection

Daily price data for both the selected asset and the market index are downloaded from Yahoo Finance and combined into a single dataset.

```{r}
asset_data <- getSymbols(
  asset_ticker,
  src = "yahoo",
  from = start_date,
  to = end_date,
  auto.assign = FALSE
)

market_data <- getSymbols(
  market_ticker,
  src = "yahoo",
  from = start_date,
  to = end_date,
  auto.assign = FALSE
)

asset_prices <- Ad(asset_data)
market_prices <- Ad(market_data)

prices <- merge(asset_prices, market_prices)

colnames(prices) <- c("asset", "market")

prices <- na.omit(prices)

head(prices)
```

## Price Series Visualization

The price series are inspected before computing returns. Since the asset and the market index are expressed on different price scales, this plot is mainly used as a quick check of the downloaded data.

```{r}
prices_df <- data.frame(
  date = index(prices),
  coredata(prices)
)

ggplot(prices_df, aes(x = date)) +
  geom_line(aes(y = asset, colour = "Asset")) +
  geom_line(aes(y = market, colour = "Market")) +
  labs(
    title = "Closing Prices",
    x = "Date",
    y = "Price",
    colour = NULL
  ) +
  theme_minimal()
```

## Return Calculation

Daily returns are computed from the closing price series and used as the main variables in the CAPM regression analysis.

```{r}
returns <- na.omit(Return.calculate(prices, method = "discrete"))

returns_df <- data.frame(
  date = index(returns),
  coredata(returns)
)

head(returns_df)
```

## Exploratory Analysis

The relationship between asset and market returns is briefly explored before estimating the regression model.

```{r}
ggplot(returns_df, aes(x = market, y = asset)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Asset vs Market Daily Returns",
    x = "Market Returns",
    y = "Asset Returns"
  ) +
  theme_minimal()
```

## CAPM Estimation

A simple linear regression is estimated using market returns as the independent variable and asset returns as the dependent variable.

```{r}
capm_model <- lm(asset ~ market, data = returns_df)

summary(capm_model)
```

## Model Evaluation

The estimated regression coefficients are inspected together with the fitted relationship between asset and market returns.

```{r}
alpha <- coef(capm_model)[["(Intercept)"]]
beta <- coef(capm_model)[["market"]]

cat("Alpha:", round(alpha, 6), "\n")
cat("Beta:", round(beta, 4), "\n")
```

```{r}
ggplot(returns_df, aes(x = market, y = asset)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, colour = "red") +
  labs(
    title = "CAPM Regression",
    x = "Market Returns",
    y = "Asset Returns"
  ) +
  theme_minimal()
```

## Discussion and Limitations

The estimated beta greater than one suggests that Apple exhibited higher sensitivity to market movements than the S&P 500 during the analyzed period.

While the simple CAPM specification captures a substantial portion of the variation in asset returns, the model remains a simplified representation of market behavior and does not account for additional risk factors or changing market conditions over time.
