---
title: "Option Pricing with Path Integrals"
author: "Rodrigo Kang"

format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-line-numbers: true
    code-overflow: wrap
    theme: cosmo
    embed-resources: true

execute:
  warning: false
  message: false

editor: visual
---

## Overview

This notebook accompanies the project **Option Pricing with Path Integrals** and contains the computational implementation supporting the methodology, experiments, and results discussed in the portfolio chapter.

The material presented here focuses on the generation of figures and tables supporting the option pricing workflow, including the Black-Scholes benchmark, standard Monte Carlo simulation, Metropolis-Hastings sampling of terminal log-prices, convergence analysis, and a simple market-data example.

## Imports and Configuration

This section initializes the main libraries used throughout the notebook.

The analysis combines analytical option pricing models, numerical simulation methods, Markov Chain Monte Carlo sampling, and historical market data obtained from Yahoo Finance. Figures and rendered tables are saved in a dedicated output folder for use in the portfolio chapter.

```{r}

Sys.setenv(LANGUAGE = "en")

library(quantmod)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gt)
```

```{r}

set.seed(42)

outputs_dir <- "1-option-pricing-with-path-integrals-figures"

dir.create(outputs_dir, showWarnings = FALSE)
```

## Black-Scholes Benchmark

The Black-Scholes model provides an analytical benchmark for European call option pricing under a set of simplifying assumptions.

```{r}

black_scholes_call <- function(S, K, T, r, sigma) {
  d1 <- (log(S / K) + (r + 0.5 * sigma^2) * T) / (sigma * sqrt(T))
  d2 <- d1 - sigma * sqrt(T)
  
  call_price <- S * pnorm(d1) - K * exp(-r * T) * pnorm(d2)
  
  return(call_price)
}
```

```{r}

S0 <- 100
K <- 100
T <- 1
r <- 0.05
sigma <- 0.20

bs_price <- black_scholes_call(
  S = S0,
  K = K,
  T = T,
  r = r,
  sigma = sigma
)

bs_price
```

```{r}

model_parameters <- tibble(
  Parameter = c(
    "Initial Asset Price",
    "Strike Price",
    "Time to Maturity",
    "Risk-Free Rate",
    "Volatility"
  ),
  Value = c(
    S0,
    K,
    T,
    r,
    sigma
  )
)

model_parameters_gt <- model_parameters |>
  gt() |>
  fmt_number(
    columns = Value,
    decimals = 4
  ) |>
  tab_header(
    title = "Model Parameters"
  )

model_parameters_gt
```

```{r}

gtsave(
  data = model_parameters_gt,
  filename = file.path(outputs_dir, "model-parameters.html")
)
```

## Path Integral Pricing

The path integral formulation motivates the option pricing problem in terms of possible asset price trajectories. In this notebook, this idea is represented computationally through simulated geometric Brownian motion paths.

```{r}

n_steps <- 50
dt <- T / n_steps

time_grid <- seq(
  from = 0,
  to = T,
  length.out = n_steps + 1
)
```

```{r}

simulate_gbm_path <- function(S0, r, sigma, T, n_steps) {
  dt <- T / n_steps
  
  path <- numeric(n_steps + 1)
  path[1] <- S0
  
  for (t in 2:(n_steps + 1)) {
    z <- rnorm(1)
    
    path[t] <- path[t - 1] * exp(
      (r - 0.5 * sigma^2) * dt +
        sigma * sqrt(dt) * z
    )
  }
  
  return(path)
}
```

```{r}

sample_path <- simulate_gbm_path(
  S0 = S0,
  r = r,
  sigma = sigma,
  T = T,
  n_steps = n_steps
)

sample_path_df <- tibble(
  Time = time_grid,
  Price = sample_path
)

sample_path_plot <- ggplot(
  sample_path_df,
  aes(x = Time, y = Price)
) +
  geom_line() +
  labs(
    title = "Simulated Geometric Brownian Motion Path",
    x = "Time",
    y = "Asset Price"
  ) +
  theme_minimal()

sample_path_plot
```

```{r}

ggsave(
  filename = file.path(outputs_dir, "simulated-gbm-path.png"),
  plot = sample_path_plot,
  width = 8,
  height = 5
)
```

## Monte Carlo Estimation

Monte Carlo simulation provides a numerical benchmark for option pricing. Multiple price trajectories are simulated, the European call payoff is evaluated at maturity, and the option price is estimated as the discounted average payoff.

```{r}

european_call_payoff <- function(ST, K) {
  payoff <- max(ST - K, 0)
  
  return(payoff)
}
```

```{r}

monte_carlo_call_price <- function(S0, K, T, r, sigma, n_steps, n_simulations) {
  payoffs <- numeric(n_simulations)
  
  for (i in 1:n_simulations) {
    path <- simulate_gbm_path(
      S0 = S0,
      r = r,
      sigma = sigma,
      T = T,
      n_steps = n_steps
    )
    
    ST <- tail(path, 1)
    
    payoffs[i] <- european_call_payoff(
      ST = ST,
      K = K
    )
  }
  
  price <- exp(-r * T) * mean(payoffs)
  
  return(price)
}
```

```{r}

n_simulations <- 10000

mc_price <- monte_carlo_call_price(
  S0 = S0,
  K = K,
  T = T,
  r = r,
  sigma = sigma,
  n_steps = n_steps,
  n_simulations = n_simulations
)

mc_price
```

## Metropolis-Hastings Estimation

The Metropolis--Hastings algorithm is used to generate samples from the target distribution implied by the terminal log-price under the Black--Scholes framework. Since the payoff of a European call depends only on the terminal asset price, sampling terminal log-prices is sufficient for this simplified setting.

```{r}

metropolis_hastings_terminal_price <- function(
    S0,
    K,
    T,
    r,
    sigma,
    n_samples,
    proposal_std
) {
  mu <- log(S0) + (r - 0.5 * sigma^2) * T
  sigma_T <- sigma * sqrt(T)
  
  current <- rnorm(
    n = 1,
    mean = mu,
    sd = sigma_T
  )
  
  payoffs <- numeric(n_samples)
  
  for (i in 1:n_samples) {
    proposal <- current + rnorm(
      n = 1,
      mean = 0,
      sd = proposal_std
    )
    
    current_density <- dnorm(
      current,
      mean = mu,
      sd = sigma_T
    )
    
    proposal_density <- dnorm(
      proposal,
      mean = mu,
      sd = sigma_T
    )
    
    acceptance_ratio <- proposal_density / current_density
    
    if (runif(1) < min(1, acceptance_ratio)) {
      current <- proposal
    }
    
    ST <- exp(current)
    
    payoffs[i] <- european_call_payoff(
      ST = ST,
      K = K
    )
  }
  
  option_price <- exp(-r * T) * mean(payoffs)
  
  return(option_price)
}
```

```{r}

mh_price <- metropolis_hastings_terminal_price(
  S0 = S0,
  K = K,
  T = T,
  r = r,
  sigma = sigma,
  n_samples = 10000,
  proposal_std = 0.15
)

mh_price
```

## Comparison of Pricing Methods

The analytical Black-Scholes solution is used as the reference value for evaluating the numerical estimates.

```{r}

comparison_df <- tibble(
  Method = c(
    "Black-Scholes",
    "Monte Carlo",
    "Metropolis-Hastings"
  ),
  Option_Price = c(
    bs_price,
    mc_price,
    mh_price
  )
) |>
  mutate(
    Absolute_Error = abs(Option_Price - bs_price)
  )

comparison_df
```

```{r}

comparison_gt <- comparison_df |>
  gt() |>
  fmt_number(
    columns = c(Option_Price, Absolute_Error),
    decimals = 4
  ) |>
  tab_header(
    title = "Comparison of Pricing Methods"
  )

comparison_gt
```

```{r}

gtsave(
  data = comparison_gt,
  filename = file.path(outputs_dir, "pricing-method-comparison.html")
)
```

```{r}

comparison_plot <- ggplot(
  comparison_df,
  aes(x = Method, y = Option_Price)
) +
  geom_col() +
  geom_hline(
    yintercept = bs_price,
    linetype = "dashed"
  ) +
  labs(
    title = "Comparison of Pricing Methods",
    x = "Method",
    y = "Option Price"
  ) +
  theme_minimal()

comparison_plot
```

```{r}

ggsave(
  filename = file.path(outputs_dir, "pricing-method-comparison.png"),
  plot = comparison_plot,
  width = 8,
  height = 5
)
```

## Convergence Analysis

The convergence analysis evaluates how the numerical estimates behave as the number of samples increases.

```{r}

sample_sizes <- c(
  100,
  500,
  1000,
  5000,
  10000
)

convergence_results <- lapply(
  sample_sizes,
  function(n_samples) {
    mc_estimate <- monte_carlo_call_price(
      S0 = S0,
      K = K,
      T = T,
      r = r,
      sigma = sigma,
      n_steps = n_steps,
      n_simulations = n_samples
    )
    
    mh_estimate <- metropolis_hastings_terminal_price(
      S0 = S0,
      K = K,
      T = T,
      r = r,
      sigma = sigma,
      n_samples = n_samples,
      proposal_std = 0.15
    )
    
    tibble(
      Samples = n_samples,
      Monte_Carlo = mc_estimate,
      Metropolis_Hastings = mh_estimate,
      Black_Scholes = bs_price
    )
  }
)

convergence_df <- bind_rows(convergence_results)

convergence_df
```

```{r}

convergence_gt <- convergence_df |>
  gt() |>
  fmt_number(
    columns = c(
      Monte_Carlo,
      Metropolis_Hastings,
      Black_Scholes
    ),
    decimals = 4
  ) |>
  tab_header(
    title = "Convergence Analysis"
  )

convergence_gt
```

```{r}

gtsave(
  data = convergence_gt,
  filename = file.path(outputs_dir, "convergence-analysis.html")
)
```

```{r}

convergence_long <- convergence_df |>
  select(
    Samples,
    Monte_Carlo,
    Metropolis_Hastings,
    Black_Scholes
  ) |>
  pivot_longer(
    cols = -Samples,
    names_to = "Method",
    values_to = "Option_Price"
  )

convergence_plot <- ggplot(
  convergence_long,
  aes(x = Samples, y = Option_Price, group = Method)
) +
  geom_line(
    aes(linetype = Method)
  ) +
  geom_point(
    aes(shape = Method)
  ) +
  scale_x_log10() +
  labs(
    title = "Convergence of Numerical Pricing Methods",
    x = "Number of Samples",
    y = "Option Price"
  ) +
  theme_minimal()

convergence_plot
```

```{r}

ggsave(
  filename = file.path(outputs_dir, "convergence-analysis.png"),
  plot = convergence_plot,
  width = 8,
  height = 5
)
```

## Market Data Example

The previous sections validated the numerical implementation under controlled assumptions. This section applies the same workflow to a simple market-based example using historical price data.

```{r}

ticker <- "AAPL"

start_date <- as.Date("2023-01-01")
end_date <- as.Date("2024-01-01")
```

```{r}

market_data <- getSymbols(
  ticker,
  src = "yahoo",
  from = start_date,
  to = end_date,
  auto.assign = FALSE
)

prices <- Ad(market_data)

colnames(prices) <- ticker

head(prices)
```

```{r}

prices_df <- tibble(
  Date = index(prices),
  Price = as.numeric(prices[, ticker])
)

market_price_plot <- ggplot(
  prices_df,
  aes(x = Date, y = Price)
) +
  geom_line() +
  labs(
    title = paste(ticker, "Adjusted Closing Prices"),
    x = "Date",
    y = "Price"
  ) +
  theme_minimal()

market_price_plot
```

```{r}

ggsave(
  filename = file.path(outputs_dir, "market-price-series.png"),
  plot = market_price_plot,
  width = 8,
  height = 5
)
```

```{r}

log_returns <- diff(log(prices))
log_returns <- na.omit(log_returns)

log_returns_df <- tibble(
  Date = index(log_returns),
  Log_Return = as.numeric(log_returns[, ticker])
)

returns_plot <- ggplot(
  log_returns_df,
  aes(x = Date, y = Log_Return)
) +
  geom_line() +
  labs(
    title = paste(ticker, "Daily Log Returns"),
    x = "Date",
    y = "Log Return"
  ) +
  theme_minimal()

returns_plot
```

```{r}

ggsave(
  filename = file.path(outputs_dir, "market-log-returns.png"),
  plot = returns_plot,
  width = 8,
  height = 5
)
```

```{r}

historical_volatility <- sd(as.numeric(log_returns)) * sqrt(252)

historical_volatility
```

```{r}

S_market <- as.numeric(last(prices))
K_market <- S_market

T_market <- 1
r_market <- 0.05
sigma_market <- historical_volatility

market_inputs <- tibble(
  Parameter = c(
    "Ticker",
    "Spot Price",
    "Strike Price",
    "Time to Maturity",
    "Risk-Free Rate",
    "Historical Volatility"
  ),
  Value = c(
    ticker,
    round(S_market, 4),
    round(K_market, 4),
    round(T_market, 4),
    round(r_market, 4),
    round(sigma_market, 4)
  )
)

market_inputs
```

```{r}

market_inputs_gt <- market_inputs |>
  gt() |>
  tab_header(
    title = "Market-Based Pricing Inputs"
  )

market_inputs_gt
```

```{r}

gtsave(
  data = market_inputs_gt,
  filename = file.path(outputs_dir, "market-pricing-inputs.html")
)
```

```{r}

bs_market_price <- black_scholes_call(
  S = S_market,
  K = K_market,
  T = T_market,
  r = r_market,
  sigma = sigma_market
)

mc_market_price <- monte_carlo_call_price(
  S0 = S_market,
  K = K_market,
  T = T_market,
  r = r_market,
  sigma = sigma_market,
  n_steps = n_steps,
  n_simulations = 10000
)

mh_market_price <- metropolis_hastings_terminal_price(
  S0 = S_market,
  K = K_market,
  T = T_market,
  r = r_market,
  sigma = sigma_market,
  n_samples = 10000,
  proposal_std = 0.15
)
```

```{r}

market_comparison <- tibble(
  Method = c(
    "Black-Scholes",
    "Monte Carlo",
    "Metropolis-Hastings"
  ),
  Option_Price = c(
    bs_market_price,
    mc_market_price,
    mh_market_price
  )
) |>
  mutate(
    Absolute_Error = abs(Option_Price - bs_market_price)
  )

market_comparison
```

```{r}

market_comparison_gt <- market_comparison |>
  gt() |>
  fmt_number(
    columns = c(Option_Price, Absolute_Error),
    decimals = 4
  ) |>
  tab_header(
    title = "Market-Based Option Pricing Comparison"
  )

market_comparison_gt
```

```{r}

gtsave(
  data = market_comparison_gt,
  filename = file.path(outputs_dir, "market-pricing-comparison.html")
)
```

```{r}

market_comparison_plot <- ggplot(
  market_comparison,
  aes(x = Method, y = Option_Price)
) +
  geom_col() +
  geom_hline(
    yintercept = bs_market_price,
    linetype = "dashed"
  ) +
  labs(
    title = "Market-Based Option Pricing Comparison",
    x = "Method",
    y = "Option Price"
  ) +
  theme_minimal()

market_comparison_plot
```

```{r}

ggsave(
  filename = file.path(outputs_dir, "market-pricing-comparison.png"),
  plot = market_comparison_plot,
  width = 8,
  height = 5
)
```

## Summary

This notebook implemented a reproducible workflow for pricing European call options using analytical and simulation-based methods.

The Black--Scholes formula was used as the analytical benchmark, standard Monte Carlo simulation provided a conventional numerical reference, and Metropolis--Hastings sampling was used to estimate prices from the terminal risk-neutral distribution.

The results show how the numerical methods can be validated against a known analytical solution and then applied to a simple market-based example using historical volatility estimated from real price data.
