---
title: "Credit Card Fraud Detection: Supervised Learning Analysis"
author: "Rodrigo Kang"

format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-line-numbers: true
    code-overflow: wrap
    theme: cosmo
    embed-resources: true

execute:
  warning: false
  message: false

editor: visual
---

## Overview

This notebook accompanies the project **Credit Card Fraud Detection** and contains the R / Quarto implementation of the supervised learning analysis.

The complete benchmark across the full set of candidate models was conducted in the Python implementation. Rather than repeating every experiment in R, this notebook focuses on the two strongest approaches retained from that benchmark:

- Random Forest trained on the original class distribution.
- XGBoost trained on a SMOTE-resampled training set.

This keeps the R implementation practical to run while preserving the main comparison that matters for the project: whether a strong tree-based ensemble can perform well under the original imbalance, and how much is gained when the minority fraud class is explicitly reinforced during training.

## Imports and Configuration

The analysis relies on a collection of packages for data manipulation, visualization, model training, performance evaluation, and class imbalance treatment. A fixed random seed is used throughout the notebook to ensure reproducibility of the results.

```{r}

# Suppress warnings
options(warn = -1)

# Data manipulation
library(dplyr)
library(readr)
library(tibble)
library(tidyr)

# Visualization
library(ggplot2)

# Tables
library(gt)

# Data splitting and preprocessing
library(rsample)
library(recipes)

# Modeling and evaluation
library(ranger)
library(xgboost)
library(yardstick)

# Resampling
library(themis)
```

```{r}
# Reproducibility
random_state <- 42
set.seed(random_state)

# Output directory
figures_dir <- "1-credit-card-fraud-detection-supervised-figures"

if (!dir.exists(figures_dir)) {
  dir.create(figures_dir, recursive = TRUE)
}
```

## Data Loading

The dataset is loaded into memory and its dimensions are inspected before the supervised modeling workflow is prepared.

```{r}
# Load dataset
df <- read_csv(
  "../data/creditcard.csv",
  show_col_types = FALSE
)

# Display dataset dimensions
cat(
  paste0(
    "Dataset shape: ",
    nrow(df),
    " rows x ",
    ncol(df),
    " columns"
  )
)
```

## Data Preparation

The target variable is converted into a factor with two levels: legitimate and fraudulent. The dataset is then split into training and testing subsets using stratified sampling, so that the rare fraud class is represented consistently in both sets.

Feature scaling is applied after the train-test split. The scaling parameters are estimated from the training data only and then applied to both the training and test sets. This avoids leaking information from the test set into the model training process.

```{r}
# Convert target variable to factor
df <- df %>%
  mutate(
    Class = factor(
      Class,
      levels = c(0, 1),
      labels = c("legitimate", "fraudulent")
    )
  )

# Train-test split
data_split <- initial_split(
  df,
  prop = 0.80,
  strata = Class
)

train_data <- training(data_split)
test_data  <- testing(data_split)

cat(
  paste0(
    "Training observations: ",
    nrow(train_data),
    "\n"
  )
)

cat(
  paste0(
    "Testing observations: ",
    nrow(test_data)
  )
)
```

```{r}
# Feature scaling recipe
scaling_recipe <- recipe(
  Class ~ .,
  data = train_data
) %>%
  step_normalize(
    all_predictors()
  )

# Estimate scaling parameters using training data only
scaling_recipe_prep <- prep(
  scaling_recipe,
  training = train_data
)

# Apply scaling to training and test sets
train_scaled <- bake(
  scaling_recipe_prep,
  new_data = train_data
)

test_scaled <- bake(
  scaling_recipe_prep,
  new_data = test_data
)
```

```{r}
# Separate predictors and target
x_train <- train_scaled %>%
  select(-Class)

y_train <- train_scaled$Class

x_test <- test_scaled %>%
  select(-Class)

y_test <- test_scaled$Class
```

```{r}
# Binary target for xgboost
y_test_binary <- ifelse(
  y_test == "fraudulent",
  1,
  0
)
```

## Class Imbalance Treatment

The original training set is kept unchanged for the Random Forest model. A second training set is created using SMOTE for XGBoost. This choice follows the Python benchmark: the R notebook is not intended to repeat the full benchmark, but to reproduce the two strongest and most informative final configurations.

```{r}
# Original class distribution
train_original <- train_scaled
```

```{r}
# SMOTE-resampled training set
smote_recipe <- recipe(
  Class ~ .,
  data = train_scaled
) %>%
  step_smote(
    Class,
    over_ratio = 1,
    seed = random_state
  )

smote_recipe_prep <- prep(
  smote_recipe,
  training = train_scaled
)

train_smote <- bake(
  smote_recipe_prep,
  new_data = NULL
)
```

```{r}
# Check class distributions after resampling
class_distribution <- bind_rows(
  train_original %>%
    count(Class) %>%
    mutate(strategy = "Original"),

  train_smote %>%
    count(Class) %>%
    mutate(strategy = "SMOTE")
) %>%
  group_by(strategy) %>%
  mutate(
    percentage = 100 * n / sum(n)
  ) %>%
  ungroup() %>%
  select(
    strategy,
    Class,
    n,
    percentage
  )

class_distribution
```

```{r}
# Save and display class distribution table
class_distribution_table <- class_distribution %>%
  gt() %>%
  tab_header(
    title = "Class Distribution by Training Strategy"
  ) %>%
  fmt_number(
    columns = percentage,
    decimals = 3
  )

gtsave(
  class_distribution_table,
  filename = file.path(
    figures_dir,
    "class-distribution-training-strategies.html"
  )
)

class_distribution_table
```

```{r}
# Helper function to create xgboost matrices
create_xgb_matrix <- function(data) {

  x <- data %>%
    select(-Class) %>%
    as.matrix()

  y <- ifelse(
    data$Class == "fraudulent",
    1,
    0
  )

  xgb.DMatrix(
    data = x,
    label = y
  )
}
```

```{r}
# Matrices for XGBoost
dtrain_smote <- create_xgb_matrix(train_smote)

dtest <- xgb.DMatrix(
  data = x_test %>% as.matrix(),
  label = y_test_binary
)
```

## Evaluation Functions

The models are evaluated using metrics that remain informative under severe class imbalance. Accuracy is deliberately not used as the main decision criterion because a classifier can achieve a high accuracy score while missing most fraudulent transactions.

The evaluation includes precision, recall, F1-score, ROC-AUC, Precision-Recall AUC, and the Kolmogorov-Smirnov statistic. Precision-Recall AUC receives particular attention because it focuses directly on the model's ability to rank the rare positive class.

```{r}
# Function to compute KS statistic
ks_statistic <- function(truth, estimate) {

  positive_scores <- estimate[truth == "fraudulent"]
  negative_scores <- estimate[truth == "legitimate"]

  ks_result <- ks.test(
    positive_scores,
    negative_scores
  )

  as.numeric(ks_result$statistic)
}
```

```{r}
# Common evaluation function
evaluate_predictions <- function(truth, predicted_class, predicted_probability) {

  evaluation_data <- tibble(
    truth = truth,
    predicted_class = predicted_class,
    fraudulent = predicted_probability
  )

  precision_value <- precision(
    evaluation_data,
    truth = truth,
    estimate = predicted_class,
    event_level = "second"
  )$.estimate

  recall_value <- recall(
    evaluation_data,
    truth = truth,
    estimate = predicted_class,
    event_level = "second"
  )$.estimate

  f1_value <- f_meas(
    evaluation_data,
    truth = truth,
    estimate = predicted_class,
    event_level = "second"
  )$.estimate

  roc_auc_value <- roc_auc(
    evaluation_data,
    truth = truth,
    fraudulent,
    event_level = "second"
  )$.estimate

  pr_auc_value <- pr_auc(
    evaluation_data,
    truth = truth,
    fraudulent,
    event_level = "second"
  )$.estimate

  ks_value <- ks_statistic(
    truth = truth,
    estimate = predicted_probability
  )

  tibble(
    precision = precision_value,
    recall = recall_value,
    f1_score = f1_value,
    roc_auc = roc_auc_value,
    pr_auc = pr_auc_value,
    ks_statistic = ks_value
  )
}
```

```{r}
# Helper function to convert probabilities into predicted classes
predict_class_from_probability <- function(probability, threshold = 0.5) {

  factor(
    ifelse(
      probability >= threshold,
      "fraudulent",
      "legitimate"
    ),
    levels = c(
      "legitimate",
      "fraudulent"
    )
  )
}
```

## Random Forest with Original Class Distribution

The first model is a Random Forest trained on the original class distribution. This configuration is useful because it tests how far a strong ensemble method can go without altering the training distribution.

Random Forest is a practical baseline for this problem. It can capture nonlinear relationships among the anonymized transaction features, is relatively robust, and does not require the same level of feature engineering as simpler linear classifiers.

```{r}
# Train Random Forest model
rf_original_model <- ranger(
  formula = Class ~ .,
  data = train_original,
  num.trees = 100,
  probability = TRUE,
  importance = "impurity",
  seed = random_state,
  num.threads = parallel::detectCores()
)
```

```{r}
# Predict probabilities on the test set
rf_original_predictions <- predict(
  rf_original_model,
  data = x_test
)$predictions

rf_original_prob <- rf_original_predictions[, "fraudulent"]
```

```{r}
# Convert probabilities to predicted classes
rf_original_class <- predict_class_from_probability(
  probability = rf_original_prob,
  threshold = 0.5
)
```

```{r}
# Evaluate Random Forest model
rf_original_results <- evaluate_predictions(
  truth = y_test,
  predicted_class = rf_original_class,
  predicted_probability = rf_original_prob
) %>%
  mutate(
    strategy = "Original",
    model = "Random Forest"
  ) %>%
  select(
    strategy,
    model,
    precision,
    recall,
    f1_score,
    roc_auc,
    pr_auc,
    ks_statistic
  )

rf_original_results
```

```{r}
# Save and display Random Forest results table
rf_original_results_table <- rf_original_results %>%
  gt() %>%
  tab_header(
    title = "Random Forest Results: Original Class Distribution"
  ) %>%
  fmt_number(
    columns = c(
      precision,
      recall,
      f1_score,
      roc_auc,
      pr_auc,
      ks_statistic
    ),
    decimals = 4
  )

gtsave(
  rf_original_results_table,
  filename = file.path(
    figures_dir,
    "random-forest-original-results.html"
  )
)

rf_original_results_table
```

## XGBoost with SMOTE

The second model is XGBoost trained on the SMOTE-resampled training set. This configuration represents the strongest imbalance-aware approach retained from the Python benchmark.

SMOTE increases the representation of fraudulent transactions by generating synthetic minority-class examples. In this setting, the goal is not to make the real-world transaction stream look balanced, but to give the model a clearer view of the regions where fraudulent transactions occur.

```{r}
# Train XGBoost model using SMOTE-resampled data
xgb_smote_model <- xgb.train(
  data = dtrain_smote,
  objective = "binary:logistic",
  eval_metric = "logloss",
  nrounds = 100,
  max_depth = 6,
  eta = 0.3,
  subsample = 1,
  colsample_bytree = 1,
  seed = random_state,
  verbose = 0
)
```

```{r}
# Predict probabilities on the test set
xgb_smote_prob <- predict(
  xgb_smote_model,
  newdata = dtest
)
```

```{r}
# Convert probabilities to predicted classes
xgb_smote_class <- predict_class_from_probability(
  probability = xgb_smote_prob,
  threshold = 0.5
)
```

```{r}
# Evaluate XGBoost with SMOTE
xgb_smote_results <- evaluate_predictions(
  truth = y_test,
  predicted_class = xgb_smote_class,
  predicted_probability = xgb_smote_prob
) %>%
  mutate(
    strategy = "SMOTE",
    model = "XGBoost"
  ) %>%
  select(
    strategy,
    model,
    precision,
    recall,
    f1_score,
    roc_auc,
    pr_auc,
    ks_statistic
  )

xgb_smote_results
```

```{r}
# Save and display XGBoost SMOTE results table
xgb_smote_results_table <- xgb_smote_results %>%
  gt() %>%
  tab_header(
    title = "XGBoost Results: SMOTE"
  ) %>%
  fmt_number(
    columns = c(
      precision,
      recall,
      f1_score,
      roc_auc,
      pr_auc,
      ks_statistic
    ),
    decimals = 4
  )

gtsave(
  xgb_smote_results_table,
  filename = file.path(
    figures_dir,
    "xgboost-smote-results.html"
  )
)

xgb_smote_results_table
```

## Final Model Comparison

The final comparison is restricted to the two retained configurations. This keeps the R notebook aligned with its purpose: reproducing the strongest final models from the broader Python benchmark without rerunning every candidate algorithm.

```{r}
# Combine final model results
final_results <- bind_rows(
  rf_original_results,
  xgb_smote_results
) %>%
  arrange(
    desc(pr_auc)
  )

final_results
```

```{r}
# Save and display final comparison table
final_results_table <- final_results %>%
  gt() %>%
  tab_header(
    title = "Final Model Comparison"
  ) %>%
  fmt_number(
    columns = c(
      precision,
      recall,
      f1_score,
      roc_auc,
      pr_auc,
      ks_statistic
    ),
    decimals = 4
  )

gtsave(
  final_results_table,
  filename = file.path(
    figures_dir,
    "final-model-comparison.html"
  )
)

final_results_table
```

Precision-Recall AUC is used as the main comparison metric because the positive class is rare. In a fraud detection setting, the key question is not whether the model is usually correct on the majority class, but whether it ranks fraudulent transactions high enough to be found.

```{r}
# Final comparison plot
final_pr_auc_plot <- final_results %>%
  mutate(
    model_strategy = paste(model, strategy, sep = " - "),
    model_strategy = reorder(model_strategy, pr_auc)
  ) %>%
  ggplot(
    aes(
      x = model_strategy,
      y = pr_auc
    )
  ) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Final Model Comparison by PR-AUC",
    x = NULL,
    y = "Precision-Recall AUC"
  ) +
  theme_minimal()

ggsave(
  filename = file.path(
    figures_dir,
    "final-model-comparison-pr-auc.png"
  ),
  plot = final_pr_auc_plot,
  width = 9,
  height = 5,
  dpi = 300
)

final_pr_auc_plot
```

## Diagnostic Plots

The following diagnostic plots compare the two retained approaches from several angles. The confusion matrices show the hard classification outcomes at the default probability threshold. The ROC and Precision-Recall curves examine ranking performance across thresholds. The Kolmogorov-Smirnov plot compares the empirical score distributions for legitimate and fraudulent transactions.

```{r}
# Prediction data used by all diagnostic plots
prediction_data <- bind_rows(
  tibble(
    model = "Random Forest (Original)",
    truth = y_test,
    predicted_class = rf_original_class,
    fraudulent = rf_original_prob
  ),

  tibble(
    model = "XGBoost (SMOTE)",
    truth = y_test,
    predicted_class = xgb_smote_class,
    fraudulent = xgb_smote_prob
  )
)
```

### Confusion Matrices

```{r}
# Confusion matrix data
confusion_matrix_data <- prediction_data %>%
  count(
    model,
    truth,
    predicted_class,
    name = "n"
  ) %>%
  complete(
    model,
    truth,
    predicted_class,
    fill = list(n = 0)
  )

confusion_matrix_plot <- confusion_matrix_data %>%
  ggplot(
    aes(
      x = predicted_class,
      y = truth,
      fill = n
    )
  ) +
  geom_tile() +
  geom_text(
    aes(label = n)
  ) +
  facet_wrap(
    ~ model,
    nrow = 1
  ) +
  labs(
    title = "Confusion Matrices",
    x = "Predicted class",
    y = "Actual class"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none"
  )

ggsave(
  filename = file.path(
    figures_dir,
    "confusion-matrices.png"
  ),
  plot = confusion_matrix_plot,
  width = 10,
  height = 4.5,
  dpi = 300
)

confusion_matrix_plot
```

### Receiver Operating Characteristic Curves

```{r}
# ROC curve data
roc_curve_data <- prediction_data %>%
  group_by(model) %>%
  roc_curve(
    truth = truth,
    fraudulent,
    event_level = "second"
  )

roc_labels <- final_results %>%
  mutate(
    label = paste0(
      model,
      " (",
      strategy,
      ") AUC = ",
      round(roc_auc, 3)
    )
  ) %>%
  transmute(
    model = paste(model, strategy, sep = " - "),
    label
  )

roc_curve_plot <- roc_curve_data %>%
  ggplot(
    aes(
      x = 1 - specificity,
      y = sensitivity,
      colour = model
    )
  ) +
  geom_path(linewidth = 1) +
  geom_abline(
    linetype = "dashed"
  ) +
  labs(
    title = "ROC Curves",
    x = "False Positive Rate",
    y = "True Positive Rate",
    colour = "Model"
  ) +
  theme_minimal()

ggsave(
  filename = file.path(
    figures_dir,
    "roc-curves.png"
  ),
  plot = roc_curve_plot,
  width = 8,
  height = 6,
  dpi = 300
)

roc_curve_plot
```

### Precision-Recall Curves

```{r}
# Precision-Recall curve data
pr_curve_data <- prediction_data %>%
  group_by(model) %>%
  pr_curve(
    truth = truth,
    fraudulent,
    event_level = "second"
  )

pr_curve_plot <- pr_curve_data %>%
  ggplot(
    aes(
      x = recall,
      y = precision,
      colour = model
    )
  ) +
  geom_path(linewidth = 1) +
  labs(
    title = "Precision-Recall Curves",
    x = "Recall",
    y = "Precision",
    colour = "Model"
  ) +
  theme_minimal()

ggsave(
  filename = file.path(
    figures_dir,
    "precision-recall-curves.png"
  ),
  plot = pr_curve_plot,
  width = 8,
  height = 6,
  dpi = 300
)

pr_curve_plot
```

### Kolmogorov-Smirnov Statistic

The Kolmogorov-Smirnov statistic measures the maximum separation between the score distributions of legitimate and fraudulent transactions. A larger value indicates stronger separation between the two classes.

```{r}
# KS summary table
ks_results <- final_results %>%
  mutate(
    model_strategy = paste(model, strategy, sep = " - ")
  ) %>%
  select(
    model_strategy,
    ks_statistic
  )

ks_results_table <- ks_results %>%
  gt() %>%
  tab_header(
    title = "Kolmogorov-Smirnov Statistic"
  ) %>%
  fmt_number(
    columns = ks_statistic,
    decimals = 4
  )

gtsave(
  ks_results_table,
  filename = file.path(
    figures_dir,
    "kolmogorov-smirnov-statistic.html"
  )
)

ks_results_table
```

```{r}
# KS empirical cumulative distribution plot
ks_plot <- prediction_data %>%
  ggplot(
    aes(
      x = fraudulent,
      colour = truth
    )
  ) +
  stat_ecdf(
    linewidth = 1
  ) +
  facet_wrap(
    ~ model,
    nrow = 1
  ) +
  labs(
    title = "Kolmogorov-Smirnov Diagnostic Plot",
    x = "Predicted probability of fraud",
    y = "Empirical cumulative distribution",
    colour = "Actual class"
  ) +
  theme_minimal()

ggsave(
  filename = file.path(
    figures_dir,
    "kolmogorov-smirnov-diagnostic-plot.png"
  ),
  plot = ks_plot,
  width = 10,
  height = 4.5,
  dpi = 300
)

ks_plot
```

## Summary

Two supervised learning strategies were retained in this R / Quarto implementation: Random Forest using the original class distribution and XGBoost with SMOTE.

This reduced benchmark is intentional. The full Python implementation already covers the broader model comparison, while the R version focuses on the final approaches worth inspecting in more detail. The results support the main empirical conclusion of the supervised stage: tree-based ensemble methods are well suited to this problem, and explicit class imbalance treatment can improve the model's ability to rank fraudulent transactions.

The diagnostic plots also show why a single metric is not enough for fraud detection. Confusion matrices describe operational outcomes at a fixed threshold, ROC curves summarise general ranking ability, Precision-Recall curves focus on the rare fraud class, and the Kolmogorov-Smirnov statistic provides an additional view of class separation in the model scores.
