---
title: "Fama-French Three-Factor Model"
author: "Rodrigo Kang"

format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-line-numbers: true
    code-overflow: wrap
    theme: cosmo
    embed-resources: true

execute:
  warning: false
  message: false

editor: visual
---

## Overview

This notebook accompanies the project **CAPM and Fama–French Models** and contains the computational implementation supporting the methodology, experiments, and results discussed in the portfolio chapter.

The material presented here focuses on practical implementation details, including data preparation, model development, training procedures, evaluation workflows, and complementary computational experiments.

## Imports and Configuration

This section initializes the main R libraries used throughout the notebook.

The analysis combines historical equity data obtained from Yahoo Finance with the Fama–French daily factor dataset developed by Kenneth French.

The workflow focuses on estimating a multifactor asset pricing model capable of explaining excess returns through market, size, and value risk factors.

```{r}
# Data manipulation
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)

# Data acquisition
library(tidyquant)

# Statistical modeling
library(broom)

# Visualization
library(ggplot2)

# Tables
library(gt)

# Output directory
output_dir <- "2-fama-french-three-factor-model-figures"

dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

# Plot theme
theme_set(theme_minimal())
```

## Asset Selection and Analysis Period

Apple Inc. (`AAPL`) is used as the target asset for the empirical analysis.

The study uses daily observations between 2020 and 2024, allowing the regression model to capture different market conditions across the selected time horizon.

```{r}
# Asset and time window
ticker <- "AAPL"

start_date <- as.Date("2020-01-01")
end_date <- as.Date("2024-12-31")
```

## Data Collection

Historical adjusted closing prices are downloaded from Yahoo Finance and later combined with the Fama–French daily three-factor dataset.

The factor dataset includes:

- the market excess return factor (`Mkt-RF`),
- the size factor (`SMB`),
- the value factor (`HML`),
- and the risk-free rate (`RF`).

These variables constitute the explanatory structure of the Fama–French three-factor specification.

```{r}
# Download adjusted stock prices
stock_data <- tq_get(
  ticker,
  from = start_date,
  to = end_date,
  get = "stock.prices"
)

head(stock_data)
```

```{r}
# Download Fama-French daily three-factor data
ff_url <- "https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/F-F_Research_Data_Factors_daily_CSV.zip"

temp_file <- tempfile(fileext = ".zip")

download.file(
  ff_url,
  temp_file,
  mode = "wb"
)

ff_file <- unzip(temp_file, list = TRUE)$Name[1]

ff_raw <- read_lines(
  unz(temp_file, ff_file)
)

start_row <- which(
  grepl("^,Mkt-RF,SMB,HML,RF", ff_raw)
) + 1

end_row <- which(
  grepl("^\\s*$", ff_raw)
) |>
  (\(x) x[x > start_row])() |>
  min() - 1

ff_factors <- read_csv(
  paste(ff_raw[start_row:end_row], collapse = "\n"),
  col_names = c(
    "date",
    "Mkt_RF",
    "SMB",
    "HML",
    "RF"
  ),
  show_col_types = FALSE
) |>
  mutate(
    date = ymd(date),
    across(
      c(Mkt_RF, SMB, HML, RF),
      ~ .x / 100
    )
  ) |>
  filter(
    date >= start_date,
    date <= end_date
  )

head(ff_factors)
```

## Return Calculation and Factor Preparation

Daily returns are computed from the adjusted closing prices and transformed into excess returns by subtracting the risk-free rate.

The factor series are converted from percentage units to decimal representation in order to maintain consistency across all variables used in the regression model.

```{r}
# Compute daily stock returns
stock_returns <- stock_data |>
  arrange(date) |>
  transmute(
    date = date,
    AAPL_Return = adjusted / lag(adjusted) - 1
  ) |>
  drop_na()

head(stock_returns)
```

## Dataset Construction

The asset returns and factor series are merged into a single modeling dataset indexed by date.

This dataset constitutes the final input used in the multifactor regression analysis.

```{r}
# Merge stock returns with Fama-French factors
model_data <- stock_returns |>
  inner_join(
    ff_factors,
    by = "date"
  ) |>
  mutate(
    AAPL_Excess_Return = AAPL_Return - RF
  )

head(model_data)
```

## Exploratory Analysis

Before estimating the regression model, the behavior of the factor series and asset excess returns is briefly inspected.

This preliminary exploration helps identify the scale and variability of the explanatory variables used in the multifactor specification.

```{r}
# Create summary table
data_summary_table <- model_data |>
  summarise(
    Observations = n(),
    Start_Date = min(date),
    End_Date = max(date),
    Mean_AAPL_Excess_Return = mean(AAPL_Excess_Return),
    SD_AAPL_Excess_Return = sd(AAPL_Excess_Return),
    Mean_Mkt_RF = mean(Mkt_RF),
    Mean_SMB = mean(SMB),
    Mean_HML = mean(HML),
    Mean_RF = mean(RF)
  )

data_summary_table
```

```{r}
# Create formatted summary table
data_summary_gt <- data_summary_table |>
  gt() |>
  tab_header(
    title = "Fama-French Modeling Dataset Summary"
  ) |>
  fmt_number(
    columns = where(is.numeric),
    decimals = 6
  )

data_summary_gt
```

```{r}
# Save formatted summary table
gtsave(
  data_summary_gt,
  file.path(
    output_dir,
    "fama_french_data_summary.html"
  )
)
```

```{r}
# Plot factor series
factor_plot <- model_data |>
  select(
    date,
    Mkt_RF,
    SMB,
    HML
  ) |>
  pivot_longer(
    cols = -date,
    names_to = "Factor",
    values_to = "Value"
  ) |>
  ggplot(
    aes(
      x = date,
      y = Value,
      color = Factor
    )
  ) +
  geom_line() +
  labs(
    title = "Fama-French Daily Factor Series",
    x = "Date",
    y = "Daily Return"
  )

factor_plot
```

```{r}
# Save factor plot
ggsave(
  filename = file.path(
    output_dir,
    "fama_french_factor_series.png"
  ),
  plot = factor_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```

## Fama-French Three-Factor Model Estimation

The regression model is estimated using ordinary least squares (OLS).

The specification includes:

- an intercept term representing the asset alpha,
- the market excess return factor,
- the size factor,
- and the value factor.

The estimated coefficients measure the sensitivity of Apple excess returns to each systematic risk factor.

```{r}
# Estimate Fama-French three-factor model
ff_model <- lm(
  AAPL_Excess_Return ~ Mkt_RF + SMB + HML,
  data = model_data
)

summary(ff_model)
```

```{r}
# Extract coefficient estimates
coefficient_table <- tidy(ff_model) |>
  mutate(
    term = recode(
      term,
      "(Intercept)" = "Alpha",
      "Mkt_RF" = "Market Factor",
      "SMB" = "Size Factor",
      "HML" = "Value Factor"
    )
  )

coefficient_table
```

```{r}
# Create formatted coefficient table
coefficient_gt <- coefficient_table |>
  gt() |>
  tab_header(
    title = "Fama-French Three-Factor Model Coefficient Estimates"
  ) |>
  fmt_number(
    columns = c(
      estimate,
      std.error,
      statistic,
      p.value
    ),
    decimals = 6
  )

coefficient_gt
```

```{r}
# Save formatted coefficient table
gtsave(
  coefficient_gt,
  file.path(
    output_dir,
    "fama_french_coefficient_estimates.html"
  )
)
```

```{r}
# Extract model-level statistics
model_summary_table <- glance(ff_model) |>
  select(
    r.squared,
    adj.r.squared,
    sigma,
    statistic,
    p.value,
    AIC,
    BIC,
    nobs
  )

model_summary_table
```

```{r}
# Create formatted model summary table
model_summary_gt <- model_summary_table |>
  gt() |>
  tab_header(
    title = "Fama-French Three-Factor Model Summary Statistics"
  ) |>
  fmt_number(
    columns = where(is.numeric),
    decimals = 6
  )

model_summary_gt
```

```{r}
# Save formatted model summary table
gtsave(
  model_summary_gt,
  file.path(
    output_dir,
    "fama_french_model_summary.html"
  )
)
```

## Model Diagnostics

The fitted values and residual series are analyzed in order to evaluate the overall behavior of the regression model.

The diagnostics provide a visual assessment of model fit and residual variability across time.

```{r}
# Add fitted values and residuals
model_data <- model_data |>
  mutate(
    Fitted_Returns = fitted(ff_model),
    Residuals = residuals(ff_model)
  )

head(model_data)
```

```{r}
# Plot observed vs fitted excess returns
observed_fitted_plot <- ggplot(
  model_data,
  aes(x = date)
) +
  geom_line(
    aes(
      y = AAPL_Excess_Return,
      color = "Observed"
    ),
    alpha = 0.7
  ) +
  geom_line(
    aes(
      y = Fitted_Returns,
      color = "Fitted"
    ),
    alpha = 0.7
  ) +
  labs(
    title = "Observed vs Fitted Excess Returns",
    x = "Date",
    y = "Excess Return",
    color = NULL
  )

observed_fitted_plot
```

```{r}
# Save observed vs fitted plot
ggsave(
  filename = file.path(
    output_dir,
    "fama_french_observed_vs_fitted_returns.png"
  ),
  plot = observed_fitted_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```

```{r}
# Plot residuals over time
residual_plot <- ggplot(
  model_data,
  aes(
    x = date,
    y = Residuals
  )
) +
  geom_line() +
  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Residuals of the Fama-French Three-Factor Model",
    x = "Date",
    y = "Residual"
  )

residual_plot
```

```{r}
# Save residual plot
ggsave(
  filename = file.path(
    output_dir,
    "fama_french_residuals.png"
  ),
  plot = residual_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```

```{r}
# Plot residual distribution
residual_distribution_plot <- ggplot(
  model_data,
  aes(x = Residuals)
) +
  geom_histogram(
    bins = 40
  ) +
  labs(
    title = "Distribution of Fama-French Model Residuals",
    x = "Residual",
    y = "Frequency"
  )

residual_distribution_plot
```

```{r}
# Save residual distribution plot
ggsave(
  filename = file.path(
    output_dir,
    "fama_french_residual_distribution.png"
  ),
  plot = residual_distribution_plot,
  width = 10,
  height = 6,
  dpi = 300
)
```

## Final Remarks

The Fama-French three-factor specification extends the CAPM framework by incorporating additional sources of systematic risk beyond the market portfolio alone.

The resulting model provides a richer description of the cross-sectional behavior of excess returns and generates the figures and tables required for the portfolio chapter discussion.
