---
title: "Customer Segmentation with RFM Analysis"
author: "Rodrigo Kang"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Overview

This notebook accompanies the project **Customer Value and Retention** and contains the R implementation of the customer segmentation workflow developed using transactional data from the AdventureWorks database.

The objective is to build customer-level RFM features, identify meaningful customer segments using K-Means clustering, and construct a prioritisation score within each segment.

This notebook focuses on the final methodology and the generation of the tables and figures used throughout the project. Exploratory analyses, model selection details, and alternative approaches are documented in the accompanying Python notebook.

## Libraries and Configuration

This section imports the libraries required for database access, data manipulation, visualisation, statistical transformation, clustering, and output generation. It also defines the output folders used to save HTML tables and PNG figures.

```{r}

Sys.setenv(LANGUAGE = "en")

library(tidyverse)
library(lubridate)
library(DBI)
library(RPostgres)
library(glue)
library(gt)
library(scales)
library(cluster)
library(factoextra)
library(clusterCrit)
library(bestNormalize)

set.seed(42)

theme_set(theme_minimal())

outputs_dir <- "2-customer-value-and-retention-rfm-figures"
tables_dir <- file.path(outputs_dir, "tables")

dir.create(outputs_dir, showWarnings = FALSE)
dir.create(tables_dir, showWarnings = FALSE)
```

## Database Connection

The AdventureWorks database is hosted locally in PostgreSQL. Connection parameters are stored in a local configuration file named `local_config.R`, which is excluded from version control.

This keeps database credentials out of the notebook while preserving a simple and reproducible workflow.

A short validation query is executed after the connection is created to confirm that the database is available.

```{r}

source("local_config.R")

con <- dbConnect(
  RPostgres::Postgres(),
  dbname = DB_NAME,
  host = DB_HOST,
  port = DB_PORT,
  user = DB_USER,
  password = DB_PASSWORD
)
```

```{r}

dbGetQuery(
  con,
  "
  SELECT
      current_database() AS database_name,
      current_user AS user_name,
      current_setting('server_encoding') AS server_encoding,
      current_setting('client_encoding') AS client_encoding;
  "
)
```

## Data Extraction

The analysis uses a consolidated transactional dataset built from the AdventureWorks sales, customer, product, and territory tables.

Each row represents one order line. This level of detail is useful because it preserves product-level information while still allowing customer-level aggregation later in the workflow.

```{r}

query <- "
SELECT
    c.customerid AS customer_id,
    soh.salesorderid AS sales_order_id,
    soh.orderdate AS order_date,
    soh.duedate AS due_date,
    soh.shipdate AS ship_date,
    soh.status AS order_status,
    soh.territoryid AS territory_id,
    st.name AS territory_name,
    st.countryregioncode AS country_region_code,
    sod.salesorderdetailid AS sales_order_detail_id,
    sod.productid AS product_id,
    p.name AS product_name,
    pc.name AS product_category,
    ps.name AS product_subcategory,
    sod.orderqty AS order_quantity,
    sod.unitprice AS unit_price,
    sod.unitpricediscount AS unit_price_discount,
    (
        sod.orderqty
        * sod.unitprice
        * (1 - sod.unitpricediscount)
    ) AS line_total
FROM sales.customer AS c
INNER JOIN sales.salesorderheader AS soh
    ON c.customerid = soh.customerid
INNER JOIN sales.salesorderdetail AS sod
    ON soh.salesorderid = sod.salesorderid
INNER JOIN production.product AS p
    ON sod.productid = p.productid
LEFT JOIN production.productsubcategory AS ps
    ON p.productsubcategoryid = ps.productsubcategoryid
LEFT JOIN production.productcategory AS pc
    ON ps.productcategoryid = pc.productcategoryid
LEFT JOIN sales.salesterritory AS st
    ON soh.territoryid = st.territoryid;
"

transactions <- dbGetQuery(con, query) %>%
  as_tibble() %>%
  mutate(
    order_date = as.Date(order_date),
    due_date = as.Date(due_date),
    ship_date = as.Date(ship_date),
    line_total = as.numeric(line_total)
  )
```

```{r}

glimpse(transactions)
```

## Initial Data Validation

Before constructing the RFM features, a compact validation table is created to confirm the size and coverage of the extracted dataset.

```{r}

inspection_summary <- tibble(
  metric = c(
    "Order Lines",
    "Customers",
    "Orders",
    "Products",
    "Territories",
    "Total Revenue"
  ),
  value = c(
    nrow(transactions),
    n_distinct(transactions$customer_id),
    n_distinct(transactions$sales_order_id),
    n_distinct(transactions$product_id),
    n_distinct(transactions$territory_id),
    round(sum(transactions$line_total, na.rm = TRUE), 2)
  )
)

inspection_summary
```

```{r}

inspection_summary_table <- inspection_summary %>%
  mutate(
    value = case_when(
      metric == "Total Revenue" ~ dollar(value, accuracy = 0.01),
      TRUE ~ comma(value)
    )
  ) %>%
  gt() %>%
  tab_header(
    title = "Initial Transaction Data Validation"
  ) %>%
  cols_label(
    metric = "Metric",
    value = "Value"
  )

inspection_summary_table
```

```{r}

gtsave(
  inspection_summary_table,
  filename = file.path(tables_dir, "rfm_initial_transaction_validation.html")
)
```

## Monthly Sales Overview

Before constructing the RFM features, a monthly summary is created to inspect the overall sales activity during the observation period.

The final incomplete month is excluded from the plot to avoid interpreting a partial month as a real drop in activity.

```{r}

monthly_sales <- transactions %>%
  mutate(order_month = floor_date(order_date, "month")) %>%
  group_by(order_month) %>%
  summarise(
    revenue = sum(line_total, na.rm = TRUE),
    orders = n_distinct(sales_order_id),
    customers = n_distinct(customer_id),
    .groups = "drop"
  )
```

```{r}

monthly_sales_table <- monthly_sales %>%
  mutate(
    revenue = dollar(revenue, accuracy = 0.01),
    orders = comma(orders),
    customers = comma(customers)
  ) %>%
  gt() %>%
  tab_header(
    title = "Monthly Sales Summary"
  ) %>%
  cols_label(
    order_month = "Month",
    revenue = "Revenue",
    orders = "Orders",
    customers = "Active Customers"
  )

monthly_sales_table
```

```{r}

gtsave(
  monthly_sales_table,
  filename = file.path(
    tables_dir,
    "monthly_sales_summary.html"
  )
)
```

```{r}

monthly_sales_plot_data <- monthly_sales %>%
  filter(order_month < as.Date("2025-06-01")) %>%
  pivot_longer(
    cols = c(revenue, orders, customers),
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c("revenue", "orders", "customers"),
      labels = c(
        "Monthly Revenue",
        "Monthly Orders",
        "Monthly Active Customers"
      )
    )
  )
```

```{r}

monthly_sales_plot <- ggplot(
  monthly_sales_plot_data,
  aes(x = order_month, y = value)
) +
  geom_line(linewidth = 1) +
  facet_grid(
    metric ~ .,
    scales = "free_y"
  ) +
  labs(
    title = "Monthly Sales Activity",
    x = NULL,
    y = NULL
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

monthly_sales_plot
```

```{r}

ggsave(
  filename = file.path(
    outputs_dir,
    "monthly_sales_activity.png"
  ),
  plot = monthly_sales_plot,
  width = 10,
  height = 8,
  dpi = 300
)
```

The dataset shows a clear increase in customer activity throughout the observation period. Revenue, order volume, and the number of active customers all exhibit sustained growth, particularly during the second half of the dataset.

This trend provides useful context for the customer segmentation analysis developed in the following sections.

## RFM Feature Construction

The RFM framework summarises customer behaviour using three dimensions:

- **Recency**: how recently a customer placed an order.
- **Frequency**: how often a customer purchased.
- **Monetary Value**: the total revenue generated by the customer.

These features provide a compact representation of customer engagement and form the basis of the segmentation model developed later in the notebook.

```{r}

reference_date <- max(transactions$order_date)

reference_date
```

The most recent transaction date in the dataset is used as the reference date for the recency calculation.

```{r}

rfm <- transactions %>%
  group_by(customer_id) %>%
  summarise(
    recency = as.numeric(
      reference_date - max(order_date)
    ),
    frequency = n_distinct(sales_order_id),
    monetary = sum(line_total),
    .groups = "drop"
  )
```

```{r}

glimpse(rfm)
```

```{r}

head(rfm)
```

The resulting dataset contains one row per customer and summarises purchasing behaviour across the entire observation period. These features form the basis of the customer segmentation model developed in the following sections.

```{r}

rfm_summary <- rfm %>%
  summarise(
    customers = n(),
    min_recency = min(recency),
    max_recency = max(recency),
    mean_recency = mean(recency),
    min_frequency = min(frequency),
    max_frequency = max(frequency),
    mean_frequency = mean(frequency),
    min_monetary = min(monetary),
    max_monetary = max(monetary),
    mean_monetary = mean(monetary)
  )
```

```{r}

rfm_summary_table <- tibble(
  Metric = c(
    "Customers",
    "Mean Recency",
    "Mean Frequency",
    "Mean Monetary Value"
  ),
  Value = c(
    comma(rfm_summary$customers),
    round(rfm_summary$mean_recency, 2),
    round(rfm_summary$mean_frequency, 2),
    dollar(rfm_summary$mean_monetary)
  )
) %>%
  gt() %>%
  tab_header(
    title = "RFM Dataset Summary"
  )

rfm_summary_table
```

```{r}

gtsave(
  rfm_summary_table,
  filename = file.path(
    tables_dir,
    "rfm_dataset_summary.html"
  )
)
```

The RFM dataset contains one observation per customer and captures purchasing behaviour across the entire observation period.

These variables will be transformed and standardised before clustering to ensure that no single dimension dominates the segmentation process.

## 

## Feature Transformation

The original RFM variables exhibit substantial differences in scale and distribution. Prior to clustering, a Box-Cox transformation is applied to reduce skewness and improve comparability across features.

Following the Python implementation, one unit is added before applying the Box-Cox transformation to keep all input values strictly positive.

The transformed variables are then standardised to ensure that each dimension contributes equally to the clustering process.

```{r}

rfm_transformed <- rfm

rfm_features <- c(
  "recency",
  "frequency",
  "monetary"
)

boxcox_lambdas <- list()

for (feature in rfm_features) {
  
  boxcox_model <- boxcox(
    rfm_transformed[[feature]] + 1
  )
  
  rfm_transformed[[paste0(feature, "_boxcox")]] <- predict(
    boxcox_model,
    rfm_transformed[[feature]] + 1
  )
  
  boxcox_lambdas[[feature]] <- boxcox_model$lambda
}

boxcox_lambdas
```

```{r}

score_columns <- c(
  "score_recency",
  "score_frequency",
  "score_monetary"
)

rfm_transformed[score_columns] <- scale(
  rfm_transformed[
    c(
      "recency_boxcox",
      "frequency_boxcox",
      "monetary_boxcox"
    )
  ]
)
```

The transformed and standardised variables are used as inputs to the clustering algorithm.

```{r}

glimpse(rfm_transformed)
```

```{r}

head(rfm_transformed)
```

The transformed dataset is now ready for clustering analysis. The next step is to evaluate candidate values of K and identify an appropriate segmentation structure for the customer base.

## Cluster Selection

Before fitting the final segmentation model, several candidate values of K are evaluated using complementary clustering diagnostics.

The analysis considers:

- Within-cluster sum of squares (WCSS)
- Silhouette Score
- Calinski-Harabasz Index
- Davies-Bouldin Index

These metrics provide different perspectives on cluster compactness and separation.

```{r}

clustering_data <- rfm_transformed %>%
  select(
    score_recency,
    score_frequency,
    score_monetary
  )
```

```{r}

set.seed(42)

k_values <- 2:10

cluster_metrics <- map_dfr(
  k_values,
  function(k) {
    
    km <- kmeans(
      clustering_data,
      centers = k,
      nstart = 20
    )
    
    sil <- silhouette(
      km$cluster,
      dist(clustering_data)
    )
    
    int_idx <- intCriteria(
      as.matrix(clustering_data),
      km$cluster,
      c(
        "Calinski_Harabasz",
        "Davies_Bouldin"
      )
    )
    
    tibble(
      k = k,
      wcss = km$tot.withinss,
      silhouette = mean(sil[, 3]),
      calinski_harabasz = int_idx$calinski_harabasz,
      davies_bouldin = int_idx$davies_bouldin
    )
  }
)
```

```{r}

cluster_metrics
```

a

```{r}

cluster_metrics_table <- cluster_metrics %>%
  gt() %>%
  cols_label(
    k = "K",
    wcss = "WCSS",
    silhouette = "Silhouette",
    calinski_harabasz = "Calinski-Harabasz",
    davies_bouldin = "Davies-Bouldin"
  ) %>%
  tab_header(
    title = "Cluster Selection Metrics"
  )

cluster_metrics_table
```

```{r}

gtsave(
  cluster_metrics_table,
  filename = file.path(
    tables_dir,
    "cluster_selection_metrics.html"
  )
)
```

The metrics table provides a quantitative comparison of alternative segmentation structures and serves as the basis for selecting the final number of clusters.

```{r}

cluster_metrics_long <- cluster_metrics %>%
  pivot_longer(
    cols = -k,
    names_to = "metric",
    values_to = "value"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c(
        "wcss",
        "silhouette",
        "davies_bouldin",
        "calinski_harabasz"
      ),
      labels = c(
        "WCSS",
        "Silhouette",
        "Davies-Bouldin",
        "Calinski-Harabasz"
      )
    )
  )
```

```{r}

cluster_metrics_plot <- ggplot(
  cluster_metrics_long,
  aes(k, value)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_grid(
    metric ~ .,
    scales = "free_y"
  ) +
  labs(
    title = "Cluster Selection Diagnostics",
    x = "Number of Clusters (K)",
    y = NULL
  ) +
  theme(
    strip.text = element_text(
      face = "bold",
      size = 11
    ),
    panel.grid.minor = element_blank()
  )

cluster_metrics_plot
```

```{r}

ggsave(
  filename = file.path(
    outputs_dir,
    "cluster_selection_diagnostics.png"
  ),
  plot = cluster_metrics_plot,
  width = 10,
  height = 12,
  dpi = 300
)
```

The diagnostics indicate that multiple values of K provide reasonable segmentation structures. Consistent with the detailed analysis conducted in the accompanying Python notebook, a final solution with nine customer segments is selected for the remainder of the analysis.

## Final Customer Segmentation

Based on the clustering diagnostics and the detailed analysis conducted in the Python notebook, the final K-Means model is fitted using nine clusters.

The resulting cluster labels are joined back to the original RFM features so that each segment can be interpreted in terms of customer behaviour.

```{r}

set.seed(42)

final_k <- 9

kmeans_final <- kmeans(
  clustering_data,
  centers = final_k,
  nstart = 20
)
```

```{r}

rfm_clustered <- rfm %>%
  mutate(
    cluster = kmeans_final$cluster
  )
```

```{r}

glimpse(rfm_clustered)
```

```{r}

head(rfm_clustered)
```

## Segment Profiles

The following table summarises the average RFM characteristics of each cluster. These profiles are used to interpret the behavioural meaning of each customer segment.

```{r}

cluster_profiles <- rfm_clustered %>%
  group_by(cluster) %>%
  summarise(
    customers = n(),
    avg_recency = mean(recency),
    avg_frequency = mean(frequency),
    avg_monetary = mean(monetary),
    total_monetary = sum(monetary),
    .groups = "drop"
  ) %>%
  arrange(cluster)
```

```{r}

cluster_profiles
```

```{r}

cluster_profiles_table <- cluster_profiles %>%
  mutate(
    customers = comma(customers),
    avg_recency = round(avg_recency, 2),
    avg_frequency = round(avg_frequency, 2),
    avg_monetary = dollar(avg_monetary),
    total_monetary = dollar(total_monetary)
  ) %>%
  gt() %>%
  cols_label(
    cluster = "Cluster",
    customers = "Customers",
    avg_recency = "Avg. Recency",
    avg_frequency = "Avg. Frequency",
    avg_monetary = "Avg. Monetary Value",
    total_monetary = "Total Monetary Value"
  ) %>%
  tab_header(
    title = "RFM Cluster Profiles"
  )

cluster_profiles_table
```

```{r}

gtsave(
  cluster_profiles_table,
  filename = file.path(
    tables_dir,
    "rfm_cluster_profiles.html"
  )
)
```
